---
title: "r-project"
author: "Riley Maher, Griffin Rubin, Katie Summers, Natalie Vadasz"
date: "Dec 8, 2020"
output: 
  html_document: 
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Group Project

## Import Packages

```{r}
library("ggplot2")
library('dplyr')
library('tidyverse')
library('geosphere')
```

## Importing Data

```{r}
# Reading in the sample CSV of rider data we made
rider_2019_sample <- read.csv('sample.csv', stringsAsFactors = TRUE)
head(rider_2019_sample)

# Reading in the weather data set
weather_data <- read.csv('NYCWeather2019.csv', stringsAsFactors = TRUE)
head(weather_data)
```

## Initial Data Summary

```{r}
# Initial summary of rider data set
str(rider_2019_sample)
summary(rider_2019_sample)
```

```{r}
# Initial summart of weather data set
str(weather_data)
summary(rider_2019_sample)
```


## Initial Data Analysis

### Rider Age

```{r}
rider_2019_sample$age <- 2019 - as.numeric(as.character(rider_2019_sample$birth.year))
rider_2019_sample <- filter(rider_2019_sample, age <= 80)
```

### Gender Split in Riders

```{r}
# Reclassifying the genders
# 0=unknown, 1=male, 2=female
rider_2019_sample$gender <- ifelse(rider_2019_sample$gender == 0, "Unkown",
                                  ifelse(rider_2019_sample$gender == 1, "Male", "Female"))

# Seeing the split of genders who rented bikes
rider_2019_sample %>%
  ggplot(aes(x=gender)) +
  geom_bar()
```

### Subscriber vs Customer for Riders

```{r}
# Seeing the split of user type who rented bikes
rider_2019_sample %>%
  ggplot(aes(x=usertype)) +
  geom_bar()
```

### Trip Duration

```{r}
# Range of all bike rides
rider_2019_sample <- filter(rider_2019_sample, tripduration <= 3000)
duration_range <- range(rider_2019_sample$tripduration, na.rm=TRUE)
duration_range

# Average length of a bike ride
duration_mean <- mean(rider_2019_sample$tripduration, na.rm=TRUE)
duration_mean

# Standard deviation of bike rides
duration_sd <- sd(rider_2019_sample$tripduration, na.rm=TRUE)
duration_sd
```

### Adjusting Dates in Data Sets

```{r}
# Creating columns of just month, day, and year
weather_data$DATE <- as.Date(weather_data$DATE, format = "%m/%d/%Y")
weather_data$Month <- format(weather_data$DATE, "%m")
weather_data$Day <- format(weather_data$DATE, "%d")
weather_data$Year <- format(weather_data$DATE, "%Y")
```

```{r}
rider_2019_sample$DATE <- as.Date(rider_2019_sample$starttime, format = "%Y-%m-%d")
rider_2019_sample$Month <- format(rider_2019_sample$Date, "%m")
rider_2019_sample$Day <- format(rider_2019_sample$Date, "%d")
rider_2019_sample$Year <- format(rider_2019_sample$Date, "%Y")
```

### Types of Weather per Month

```{r}
# Average precipitation per month
weather_data %>% 
  summarise(average_precip = tapply(PRCP, Month, mean, na.rm=TRUE))
```

```{r}
# Average snow per month
weather_data %>% 
  summarise(avg_snow = tapply(SNOW, Month, mean, na.rm=TRUE))
```

```{r}
# Average wind speed per month
weather_data %>%
  summarise(average_wind_speed = tapply(AWND, Month, mean, na.rm=TRUE))
```

## Exploratory Data Analysis - Weather Effects

# NOTE: I left start.station.id in as a means to check that the merge operated properly and didn't just repeat the incorrect data. Compare this to the other data frames we've created to see if the data matches up. I have, and it looks ok, but another set of fresh eyes should also check. --Riley

```{r}
# Combining data frames to compare data
edited_weather <- select(weather_data, PRCP, SNOW, AWND, DATE)
edited_rider <- select(rider_2019_sample, start.station.id, tripduration, DATE)

total_data = merge(edited_weather, edited_rider, by.x="DATE", by.y="DATE", all.x=TRUE)
head(total_data)
```

## Exploratory Data Analysis - Ride History

### Distance Between Stations

```{r}
# Distance between start and end station in Meters

rider_2019_sample <- mutate(rider_2019_sample, 
                            distance = distHaversine(cbind(rider_2019_sample$start.station.longitude,
                                                           rider_2019_sample$start.station.latitude),
                                                     cbind(rider_2019_sample$end.station.longitude,
                                                           rider_2019_sample$end.station.latitude)))

head(rider_2019_sample)
```

```{r}
# Speed of the rider

rider_2019_sample$speed <- rider_2019_sample$distance/rider_2019_sample$tripduration

# Average speed of all riders
rider_2019_sample %>%
  summarise(average_speed = mean(speed))

# Average speed of young riders
rider_2019_sample %>%
  filter(age <= 45) %>%
  summarise(young_average = mean(speed))

# Average speed of old riders
rider_2019_sample %>%
  filter(age >= 65) %>%
  summarise(old_average = mean(speed))

# Average speed of female riders
rider_2019_sample %>%
  filter(gender == "Female") %>%
  summarise(female_average = mean(speed))

# Average speed of male riders
rider_2019_sample %>%
  filter(gender == "Male") %>%
  summarise(male_average = mean(speed))
```

```{r}
# Scatter Plot of speed by age

rider_2019_sample %>%
  ggplot(aes(x = speed, y = age)) +
  geom_point(alpha = .25, color = 'blue', size = 1) +
  geom_point(shape = 1, size = 1, colour = "black") +
  labs(title="Average Speed of Riders by Age", x="Speed", y="Age")
```
big_rain_data <- total_data %>% filter(total_data$PRCP > (0.1054468 + 0.2692807))
Rain_Locations <- subset(big_rain_data, select=c("start.station.longitude", "start.station.latitude", "start.station.name"))
names(Rain_Locations) = c("long", "lat", "station")
count_starts = as.data.frame(table(big_rain_data$start.station.name), na.rm=TRUE)
names(count_starts) = c("station", "starts")
count_begins = as.data.frame(merge(count_starts,Rain_Locations, by.x = "station", by.y = "station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
count_ends = as.data.frame(table(big_rain_data$end.station.name), na.rm=TRUE)
names(count_ends) = c("station", "ends")
station_flow = as.data.frame(merge(count_begins, count_ends, by.x="station", by.y="station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
station_flow$net = station_flow$starts / station_flow$ends
station_flow = na.omit(station_flow)
distinct_data <- station_flow[!(duplicated(station_flow$station)), ]
distinct_data

#Asymmetric Travel by Station, All rides where PRCP was 1 SD above the Mean (>0.3747275)
distinct_data %>% mutate(station = fct_reorder(station, net)) %>% ggplot(aes(x=station, y=net)) + geom_bar(stat = "identity")+ geom_hline(yintercept=1, linetype="dashed", color = "red") + labs(x="Stations", y="Total Starts / Total Ends in 2019", title = "Net 2019 Flow of Bikes per Station") + theme(axis.text.x = element_blank())

```
###Average Precipitation by Age
```{r}
###Mean PRCP by Age of Rider
Plotdata <- total_data %>% 
  group_by(age) %>%
  summarise(mean_PRCP_by_age = mean(PRCP),
            mean_duration = mean(tripduration)) 
Plotdata
  ggplot(Plotdata, aes(x = age, y = mean_PRCP_by_age)) + geom_point(alpha =0.9, shape = 18, colour = "blue", size = Plotdata$mean_duration/150) + geom_smooth(colour = "orange") 
```
###Mean Wind by Age of Rider
```{r}
Plotdata2 <- total_data %>% 
  group_by(age) %>%
   summarise(mean_AWND_by_age = mean(AWND,na.rm = TRUE),
            mean_duration = mean(tripduration)) 
Plotdata2
  ggplot(Plotdata2, aes(x = age, y = mean_AWND_by_age)) + geom_point(alpha =0.9, shape = 18, colour = "cyan3", size = Plotdata$mean_duration/150) + geom_smooth(colour = "orange") 
```
###Mean Snow by Age of Rider
```{r}
Plotdata3 <- total_data %>% 
  group_by(age) %>%
   summarise(mean_SNOW_by_age = mean(SNOW,na.rm = TRUE),
            mean_duration = mean(tripduration)) 
Plotdata3
  ggplot(Plotdata3, aes(x = age, y = mean_SNOW_by_age)) + geom_point(alpha =0.9, shape = 18, colour = "deepskyblue", size = Plotdata$mean_duration/150) + geom_smooth(colour = "orange") 
```
         

#Station locations with Net Asymmetric Traffic <0.75, with PRCP >0.3747275
```{r}
Extra_Bikes_Rain <- distinct_data %>% filter(distinct_data$net < 0.75)
Lost_Bikes_Rain <- distinct_data %>% filter(distinct_data$net > 1.25)
Extra_Bikes_Rain
Lost_Bikes_Rain
Bikes_Rain_Map <- ggmap(start, extent = "device", legend = "topright")

##Location of Stations with Significantly Positive Asymmetric Traffic during High PRCP days
#below 0.75 Asymmetric Traffic
Bikes_Rain_Map + geom_point(data = Extra_Bikes_Rain, aes(x = Extra_Bikes_Rain$long, y = Extra_Bikes_Rain$lat), colour = "deepskyblue")

##Location of Stations with over 1.25 Asymmetric Traffic during High PRCP days
#More bikes taken than returned in high PRCP
Bikes_Rain_Map + geom_point(data = Lost_Bikes_Rain, aes(x = Lost_Bikes_Rain$long, y = Lost_Bikes_Rain$lat), colour = "deepskyblue")
```
#Station locations with Net Asymmetric Traffic <0.75, with SNOW > 0.21624501
#Station locations with Net Asymmetric Traffic >1.25, with SNOW > 0.21624501
```{r}
big_snow_data <- total_data %>% filter(total_data$SNOW > (0.21624501))
Snow_Locations <- subset(big_snow_data, select=c("start.station.longitude", "start.station.latitude", "start.station.name"))
names(Snow_Locations) = c("long", "lat", "station")
count_starts = as.data.frame(table(big_snow_data$start.station.name), na.rm=TRUE)
names(count_starts) = c("station", "starts")
count_begins = as.data.frame(merge(count_starts,Snow_Locations, by.x = "station", by.y = "station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
count_ends = as.data.frame(table(big_snow_data$end.station.name), na.rm=TRUE)
names(count_ends) = c("station", "ends")
station_flow = as.data.frame(merge(count_begins, count_ends, by.x="station", by.y="station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
station_flow$net = station_flow$starts / station_flow$ends
station_flow = na.omit(station_flow)
distinct_data <- station_flow[!(duplicated(station_flow$station)), ]
distinct_data

Extra_Bikes_Snow <- distinct_data %>% filter(distinct_data$net < 0.75)
Lost_Bikes_Snow <- distinct_data %>% filter(distinct_data$net > 1.25)
Bikes_Snow_Map <- ggmap(start, extent = "device", legend = "topright")

##Location of Stations with Significantly Positive Asymmetric Traffic during High Snow days
#below 0.75 Asymmetric Traffic
Bikes_Snow_Map + geom_point(data = Extra_Bikes_Snow, aes(x = Extra_Bikes_Snow$long, y = Extra_Bikes_Snow$lat), colour = "cyan")

##Location of Stations with over 1.25 Asymmetric Traffic during High Snow days
#More bikes taken than returned in high Snow
Bikes_Snow_Map + geom_point(data = Lost_Bikes_Snow, aes(x = Lost_Bikes_Snow$long, y = Lost_Bikes_Snow$lat), colour = "cyan")
```

#Station locations with Net Asymmetric Traffic <0.75, with AWND > 6.383933
#Station locations with Net Asymmetric Traffic >1.25, with AWND > 6.383933
```{r}
big_wind_data <- total_data %>% filter(total_data$AWND > (6.383933))
Wind_Locations <- subset(big_wind_data, select=c("start.station.longitude", "start.station.latitude", "start.station.name"))
names(Wind_Locations) = c("long", "lat", "station")
count_starts = as.data.frame(table(big_wind_data$start.station.name), na.rm=TRUE)
names(count_starts) = c("station", "starts")
count_begins = as.data.frame(merge(count_starts,Wind_Locations, by.x = "station", by.y = "station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
count_ends = as.data.frame(table(big_wind_data$end.station.name), na.rm=TRUE)
names(count_ends) = c("station", "ends")
station_flow = as.data.frame(merge(count_begins, count_ends, by.x="station", by.y="station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
station_flow$net = station_flow$starts / station_flow$ends
station_flow = na.omit(station_flow)
distinct_data <- station_flow[!(duplicated(station_flow$station)), ]
distinct_data

Extra_Bikes_Wind <- distinct_data %>% filter(distinct_data$net < 0.75)
Lost_Bikes_Wind <- distinct_data %>% filter(distinct_data$net > 1.25)
Bikes_Wind_Map <- ggmap(start, extent = "device", legend = "topright")

##Location of Stations with Significantly Positive Asymmetric Traffic during High Snow days
#below 0.75 Asymmetric Traffic
Bikes_Wind_Map + geom_point(data = Extra_Bikes_Wind, aes(x = Extra_Bikes_Wind$long, y = Extra_Bikes_Wind$lat), colour = "darkorchid1")

##Location of Stations with over 1.25 Asymmetric Traffic during High Snow days
#More bikes taken than returned in high Snow
Bikes_Wind_Map + geom_point(data = Lost_Bikes_Wind, aes(x = Lost_Bikes_Wind$long, y = Lost_Bikes_Wind$lat), colour = "darkorchid1")
```
#Station locations with Net Asymmetric Traffic <0.75, with TripDuration > 13026.7
#Station locations with Net Asymmetric Traffic <0.75, with TripDuration > 13026.7
```{r}
big_trip_data <- total_data %>% filter(total_data$tripduration > (13026.7))
Longtrip_Locations <- subset(big_trip_data, select=c("start.station.longitude", "start.station.latitude", "start.station.name"))
names(Longtrip_Locations) = c("long", "lat", "station")
count_starts = as.data.frame(table(big_trip_data$start.station.name), na.rm=TRUE)
names(count_starts) = c("station", "starts")
count_begins = as.data.frame(merge(count_starts,Longtrip_Locations, by.x = "station", by.y = "station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
count_ends = as.data.frame(table(big_trip_data$end.station.name), na.rm=TRUE)
names(count_ends) = c("station", "ends")
station_flow = as.data.frame(merge(count_begins, count_ends, by.x="station", by.y="station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
station_flow$net = station_flow$starts / station_flow$ends
station_flow = na.omit(station_flow)
distinct_data <- station_flow[!(duplicated(station_flow$station)), ]
distinct_data

Extra_Bikes_Longtrip <- distinct_data %>% filter(distinct_data$net < 0.75)
Lost_Bikes_Longtrip <- distinct_data %>% filter(distinct_data$net > 1.25)
Bikes_Longtrip_Map <- ggmap(start, extent = "device", legend = "topright")

##Location of Stations with Significantly Positive Asymmetric Traffic for trips lasting longer than 1 SD above the mean
#below 0.75 Asymmetric Traffic
Bikes_Longtrip_Map + geom_point(data = Extra_Bikes_Longtrip, aes(x = Extra_Bikes_Longtrip$long, y = Extra_Bikes_Longtrip$lat), colour ="darkolivegreen3")

##Location of Stations with over 1.25 Asymmetric Traffic during longer trips
#More bikes taken than returned in high Snow
Bikes_Longtrip_Map + geom_point(data = Lost_Bikes_Longtrip, aes(x = Lost_Bikes_Longtrip$long, y = Lost_Bikes_Longtrip$lat), colour ="darkolivegreen3")
```
